### 2020-05-22 serial.tex
ok
minor tweaks

Generation intervals and serial intervals are critical quantities for characterizing outbreak dynamics.
Generation intervals characterize the time between infection and transmission; whereas serial intervals characterize the time between the onset of symptoms in a chain of transmission.
They are often used interchangeably, leading to misunderstanding of how to use either generation or serial intervals to establish a link between the epidemic growth rate $r$ and the reproduction number \RR.
On one hand, generation intervals provide a mechanistic link between $r$ and \RR but are harder to measure via contact tracing.
On the other hand, serial intervals are easier to measure from contact tracing but recent studies suggest that two intervals give different estimates of \RR from $r$.
We present a general framework for characterizing \emph{any} epidemiological delays based on cohorts (i.e., a group of individuals that share the same event time, such as symptom onset) and show that \emph{forward-looking} serial intervals, which correctly estimates \RR from $r$, are not the same as ``intrinsic'' serial intervals, but instead change with $r$.
We provide a heuristic method for addressing potential biases that can arise from not accounting for changes in serial intervals across cohorts and apply the method to estimating \RR for the COVID-19 outbreak in China using serial interval data --- our analysis shows that using incorrectly defined serial intervals can severely bias the estimate.
This study demonstrates the importance of early epidemiological investigation through contact tracing and provides rationale for reassessing generation intervals, serial intervals, and \RR\ estimates, for COVID-19.

@@

The reproduction number \RR is one of the most important characteristics of an emerging epidemic, including characterization of coronavirus disease 2019 (COVID-19) \citep{majumder2020early}.
The reproduction number is defined as the average number of secondary cases caused by a primary case;
the reproduction number in a fully susceptible population --- referred to as the basic reproduction number \Ro --- allows us to predict the extent to which a disease will spread in the population and the amount of intervention necessary to eliminate disease \citep{anderson1991infectious}.
Near the beginning of an outbreak, \RR is often estimated from the observed exponential growth rate using generation- and serial-interval distributions (e.g., \cite{wallinga2007generation, fraser2009pandemic, hampson2009transmission, chunara2012social, chowell2014west, du2020serial, jung2020real}).

@@

Consider a Susceptible-Exposed-Infectious-Recovered model:
\begin{equation}
\begin{aligned}
\frac{\mathrm{d}S}{\mathrm{d}t} &= - \beta S I\\
\frac{\mathrm{d}E}{\mathrm{d}t} &= \beta S I - \sigma E\\
\frac{\mathrm{d}I}{\mathrm{d}t} &= \sigma E - \gamma I \\
\frac{\mathrm{d}R}{\mathrm{d}t} &= \gamma I \\
\end{aligned}
\end{equation}
where $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, and $1/\gamma$ is the mean infectious period.
We further assume that the latent period is equivalent to incubation period; in other words, infected individuals can only transmit after symptom onset.
Then, the generation interval will be always longer than the incubation period.

The joint probability distribution of the intrinsic incubation periods and intrinsic generation intervals for this model can be written as:
\begin{equation}
h(x, \tau) = \begin{cases}
0 & x > \tau\\
\gamma \sigma \exp(-\gamma (\tau-x)-\sigma x) & x \leq \tau
\end{cases}
\end{equation}
Then, the intrinsic generation-interval distribution is given by:
\begin{equation}
\begin{aligned}
g(\tau) &= \int_0^\tau h(x, \tau) \dx\\
&= \frac{\gamma \sigma}{\sigma-\gamma} (\exp(-\gamma \tau) - \exp(-\sigma \tau))
\end{aligned}
\end{equation}
On the other hand, the initial forward serial-interval distribution is given by:
\begin{equation}
\begin{aligned}
f_{0}(\tau) &\propto \int_{-\infty}^{0} \int_{0}^{\tau} \exp(r \pinf) h(-\pinf, \sinf - \pinf) \idist(\tau - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf\\
&\propto \int_{-\infty}^{0} \int_{0}^{\tau}\exp(r \pinf) \exp(-\gamma \sinf+ \sigma \pinf) \exp(-\sigma(\tau-\sinf)) \, \mathrm{d}\sinf\,\mathrm{d}\pinf\\
&\propto  \exp(-\sigma \tau) \int_{-\infty}^{0} \int_{0}^{\tau} \exp((\sigma-\gamma) \sinf) \exp((r+ \sigma) \pinf)\, \mathrm{d}\sinf\,\mathrm{d}\pinf\\
&\propto  (\exp(-\gamma \tau) - \exp(-\sigma \tau)) \int_{-\infty}^{0} \exp((r+ \sigma) \pinf)\,\mathrm{d}\pinf\\
&\propto  \exp(-\gamma \tau) - \exp(-\sigma \tau)
\end{aligned}
\end{equation}
Therefore, both the intrinsic generation intervals and the initial forward serial intervals are identically distributed and have the same mean.

>>>>

Generation intervals and serial intervals are critical quantities for characterizing outbreak dynamics.
Generation intervals characterize the time between infection and transmission; whereas serial intervals characterize the time between the onset of symptoms in a chain of transmission.
They are often used interchangeably, leading to misunderstanding of how to use either generation or serial intervals to establish a link between the epidemic growth rate $r$ and the reproduction number \RR.
On one hand, generation intervals provide a mechanistic link between $r$ and \RR but are harder to measure via contact tracing.
On the other hand, serial intervals are easier to measure from contact tracing but recent studies suggest that the two intervals give different estimates of \RR from $r$.
We present a general framework for characterizing \emph{any} epidemiological delays based on cohorts (i.e., a group of individuals that share the same event time, such as symptom onset) and show that \emph{forward-looking} serial intervals, which correctly estimates \RR from $r$, are not the same as ``intrinsic'' serial intervals, but instead change with $r$.
We provide a heuristic method for addressing potential biases that can arise from not accounting for changes in serial intervals across cohorts and apply the method to estimating \RR for the COVID-19 outbreak in China using serial interval data --- our analysis shows that using incorrectly defined serial intervals can severely bias the estimate.
This study demonstrates the importance of early epidemiological investigation through contact tracing and provides a rationale for reassessing generation intervals, serial intervals, and \RR\ estimates, for COVID-19.

@@

The reproduction number \RR is one of the most important characteristics of an emerging epidemic, including characterization of coronavirus disease 2019 (COVID-19) \citep{majumder2020early}.
The reproduction number is defined as the average number of secondary cases caused by a primary case;
the value in a fully susceptible population --- referred to as the basic reproduction number \Ro --- allows us to predict the extent to which an infection will spread in the population and the amount of intervention necessary to eliminate it \citep{anderson1991infectious}.
Near the beginning of an outbreak, \RR is often estimated from the observed exponential growth rate using generation- and serial-interval distributions (e.g., \cite{wallinga2007generation, fraser2009pandemic, hampson2009transmission, chunara2012social, chowell2014west, du2020serial, jung2020real}).

@@

Consider a Susceptible-Exposed-Infectious-Recovered model:
\begin{equation}
\begin{aligned}
\frac{\mathrm{d}S}{\mathrm{d}t} &= - \beta S I\\
\frac{\mathrm{d}E}{\mathrm{d}t} &= \beta S I - \gamma_E E\\
\frac{\mathrm{d}I}{\mathrm{d}t} &= \gamma_E E - \gamma_I I \\
\frac{\mathrm{d}R}{\mathrm{d}t} &= \gamma_I I \\
\end{aligned}
\end{equation}
where $\beta$ is the transmission rate, $1/\gamma_E$ is the mean latent period, and $1/\gamma_I$ is the mean infectious period.
We further assume that the latent period is equivalent to incubation period; in other words, infected individuals can only transmit after symptom onset.
Then, the generation interval will be always longer than the incubation period.

The joint probability distribution of the intrinsic incubation periods and intrinsic generation intervals for this model can be written as:
\begin{equation}
h(x, \tau) = \begin{cases}
0 & x > \tau\\
\gamma_I \gamma_E \exp(-\gamma_I (\tau-x)-\gamma_E x) & x \leq \tau
\end{cases}
\end{equation}
Then, the intrinsic generation-interval distribution is given by:
\begin{equation}
\begin{aligned}
g(\tau) &= \int_0^\tau h(x, \tau) \dx\\
&= \frac{\gamma_I \gamma_E}{\gamma_E-\gamma_I} (\exp(-\gamma_I \tau) - \exp(-\gamma_E \tau))
\end{aligned}
\end{equation}
On the other hand, the initial forward serial-interval distribution is given by:
\begin{equation}
\begin{aligned}
f_{0}(\tau) &\propto \int_{-\infty}^{0} \int_{0}^{\tau} \exp(r \pinf) h(-\pinf, \sinf - \pinf) \idist(\tau - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf\\
&\propto \int_{-\infty}^{0} \int_{0}^{\tau}\exp(r \pinf) \exp(-\gamma_I \sinf+ \gamma_E \pinf) \exp(-\gamma_E(\tau-\sinf)) \, \mathrm{d}\sinf\,\mathrm{d}\pinf\\
&\propto  \exp(-\gamma_E \tau) \int_{-\infty}^{0} \int_{0}^{\tau} \exp((\gamma_E-\gamma_I) \sinf) \exp((r+ \gamma_E) \pinf)\, \mathrm{d}\sinf\,\mathrm{d}\pinf\\
&\propto  (\exp(-\gamma_I \tau) - \exp(-\gamma_E \tau)) \int_{-\infty}^{0} \exp((r+ \gamma_E) \pinf)\,\mathrm{d}\pinf\\
&\propto  \exp(-\gamma_I \tau) - \exp(-\gamma_E \tau)
\end{aligned}
\end{equation}
Therefore, both the intrinsic generation intervals and the initial forward serial intervals are identically distributed and have the same mean.
